---
title: "The perceptron"
---

The perceptron is a fundamental concept in artificial intelligence and machine learning. It's a type of artificial neural network model introduced by **Frank Rosenblatt** in **1958**. At its core, the perceptron mimics how biological neurons function, using mathematical computations to classify data points as belonging to one of two categories.It is viewed as one of the earliest models for supervised learning in neural networks. Despite its simplicity, it laid the groundwork for more complex networks like multi-layer perceptrons, which can solve non-linear problems. While the perceptron itself only works well with linearly separable data, its advancements contributed significantly to the evolution of AI systems.

Think of the perceptron as a simple decision-making tool. It takes input signals (features of the data), applies weights to them, sums them up, and passes the result through an activation function to produce an output. If the sum crosses a certain threshold, the perceptron "fires," producing an output of 1; otherwise, it outputs 0. This makes it ideal for binary classification tasks.

Well, in order to create an artificial neural network from scratch with R, we're going to start with the perceptron, while recalling some important mathematical notions. 

Let's start with an example.

Let's suppose that we have two kind of plant that we have to divide into two classes: **toxic** plants represented by class **0** and **healthy** plants represented by class **1**. Each plant has his characteristics **X1** and **X2**. The following table show us what it is about:

| Y  |X1 | X2|
|----|---|---|
| 1  |0.5| 2 |
| 1  |1.1|2.1|
| 1  |0.7|2.6|
| 1  |2  | 1 |
| 1  |2.5|0.7|
| 1  |2.2|0.3|

Let' us have a better look with a graphic


```{r}
library(ggplot2)
x1= c(0.5,1.1,0.7,2,2.5,2.2)
x2=c(2,2.1,2.6,1,0.7,0.3)
y= factor(c(1,1,1,0,0,0), levels=c(0,1))
Y=factor(y)
data= data.frame(cbind(y,x1,x2)  ) 
ggplot(data = data, aes(x = x1, y = x2)) +
  geom_point(aes(colour = Y), size = 3) + 
  scale_colour_manual(values = c("red", "green")) 
  theme_minimal()
```

As we can see, we can provide a line that can separate the two kinds of plant. This line is **the decision boundary**. So we're able to develop a model, based on this line to predict the class of a new plant.
```{r }
library(ggplot2)
x1= c(0.5,1.1,0.7,2,2.5,2.2,1.5)
x2=c(2,2.1,2.6,1,0.7,0.3,0.5)
y= factor(c(1,1,1,0,0,0,2), levels=c(0,1,2))
Y=factor(y)
data= data.frame(cbind(y,x1,x2)  ) 
ggplot(data = data, aes(x = x1, y = x2)) +
  geom_point(aes(colour = Y), size = 3) + 
  scale_colour_manual(values = c("red", "green", "blue")) + 
  geom_abline(slope = 1, intercept = 0.1, colour = "black", size = 1.2) + # Boundary line
  labs(title = "Data separation", x = "x1", y = "x2") + 
  theme_minimal()
```


To de velop the model, we need to know the equation of the line that separate accuratly the two plants. To achieve this, we'll develop a linear model by sending **X1** and **X2** to a neuron, adding a coefficient **W**, respectively **W1** and **W2** and what we call the bias **b**. It'll lead us to :

$$  z(X1,X2)= W1*X1 + W2*X2 + b   $$


