---
title: "Hands-on"
---

Let's start with an example.

Let's suppose that we have two kind of plant that we have to divide into two classes: **toxic** plants represented by class **0** and **healthy** plants represented by class **1**. Each plant has his characteristics **X1** and **X2**. The following table show us what it is about:

| Y  |$(X_1)$ | $(X_2)$|
|----|---|---|
| 1  |0.5| 2 |
| 1  |1.1|2.1|
| 1  |0.7|2.6|
| 0  |2  | 1 |
| 0  |2.5|0.7|
| 0  |2.2|0.3|

Let' us have a better look with a graphic


```{r , warning=FALSE, message=FALSE}
library(ggplot2)
x1= c(0.5,1.1,0.7,2,2.5,2.2)
x2=c(2,2.1,2.6,1,0.7,0.3)
y= factor(c(1,1,1,0,0,0), levels=c(0,1))
Y=factor(y)
data= data.frame(cbind(y,x1,x2)  ) 
ggplot(data = data, aes(x = x1, y = x2)) +
  geom_point(aes(colour = Y), size = 3) + 
  scale_colour_manual(values = c("red", "green")) +
  theme_minimal()
```

As we can see, we can provide a line that can separate the two kinds of plant. This line is **the decision boundary**. So we're able to develop a model, based on this line to predict the class of a new plant.
```{r , warning=FALSE, message=FALSE }
library(ggplot2)
x1= c(0.5,1.1,0.7,2,2.5,2.2,1.5)
x2=c(2,2.1,2.6,1,0.7,0.3,0.5)
y= factor(c(1,1,1,0,0,0,2), levels=c(0,1,2))
Y=factor(y)
data= data.frame(cbind(y,x1,x2)  ) 
ggplot(data = data, aes(x = x1, y = x2)) +
  geom_point(aes(colour = Y), size = 3) + 
  scale_colour_manual(values = c("red", "green", "blue")) + 
  geom_abline(slope = 0.9, intercept = 0.1, colour = "black", size = 1.2) + # Boundary line
  labs(title = "Data separation", x = "x1", y = "x2") + 
  theme_minimal()
```

### The model

To develop the model, we need to know the equation of the line that separate accuratly the two plants. To achieve this, we'll develop a linear model by sending **X1** and **X2** to a neuron, adding a coefficient **W**, respectively **W1** and **W2** and what we call the bias **b**. It'll lead us to :

$$  z(X1,X2)= W1*X1 + W2*X2 + b   $$



The boundary line is the value for which $Z(X_1,X_2)=0$   So by posing $Z(X_1,X_2)=0$, we can find the expression of the equation of the boundary line.
To predict a class for a new  plant, we have to  set the parameters $W_1$, $W_2$  and  $b$  to separe in the best way the two class and after that we will be able to send a new plant according to the sign of the $Z$ computed. 
Now, to improve this model, one of the best thing that we can do is to add probabilities. The further the plant is from the line, the greater the chance that it belongs to this class. In this context, we use an **activation function** that returns a value close to 0 or 1 as we move away from the line. Generally we use the sigmoiid fonction.

### The sigmoid

The **sigmoid function** is one of the most common activation functions in artificial neural networks, especially in simpler or shallow models. It maps any real-valued number to a range between **0 and 1**, making it particularly useful for binary classification problems.


The **sigmoid function** $a(Z)$ is defined as:

$a(Z) = \frac{1}{1 + e^{-Z}}$



Where:

- $e$ is the base of the natural logarithm (approximately 2.718).
-  $Z$ is the input to the function, which can be any real number.



The graph of the sigmoid function has an **S-shape** (sigmoid curve) and is asymptotic to both 0 and 1 for extreme values of \( x \). 
```{r , warning=FALSE, message=FALSE}
library(ggplot2)
# Define sigmoid function
sigmoid <- function(x) {
  return(1 / (1 + exp(-x)))
}
x_values <- seq(-10, 10, by = 0.1)
y_values <- sigmoid(x_values)
# Create a data frame for plotting
data <- data.frame(x = x_values, y = y_values)

# Plot the sigmoid curve with color differentiation for both the curve and area under it
ggplot(data, aes(x = x, y = y)) +
  geom_line(size = 1.5, color = "black") +  # Black line for sigmoid curve
  geom_ribbon(data = subset(data, x < 5), aes(ymin = y, ymax = 1), fill = "lightblue", alpha = 0.5) +  # Light blue for the left part under and on the curve
  geom_ribbon(data = subset(data, x >=-5 ), aes(ymin = 0, ymax = y), fill = "lightcoral", alpha = 0.5) +  # Light red for the right part under the curve
  labs(title = "Sigmoid Function ", x = "x", y = "Sigmoid(x)") +
  theme_minimal() +
  theme(axis.title = element_text(size = 12), plot.title = element_text(size = 14, hjust = 0.5))
```


### The Cost Function

To find the best model, we need to minimize the error between the predictions and the real values. This process involves using the sigmoid function to derive probabilities, leveraging the Bernoulli probability function, and defining the likelihood and log-likelihood functions, that we call the log-loss function which is defined as
$$\ell(W_1, W_2, b) = -1/n[\sum_{i=1}^n \left[ Y_i \cdot \log(a(z_i)) + (1 - Y_i) \cdot \log(1 - a(z_i)) \right]]$$

Let us have a better view on how we got this.
The likelihood function measures how well the model predicts the entire dataset. To find it, we can just make the product of all the probabilities that we compute using the sigmoid function for each plants.
 For \( n \) observations, it is defined as:

$$L(W_1, W_2, b) = \prod_{i=1}^n P(Y_i|X_i)$$

S

$$P(Y_i|X_i)$$

The sigmoid function transforms a linear combination of features into probabilities, defined as:
$a(z) = \frac{1}{1 + e^{-z}}$


where:

$z = W_1 \cdot X_1 + W_2 \cdot X_2 + b$

Here,  $W_1$ , $W_2$, and $b$  are the parameters of the model, and  $X_1$,  $X_2$  are the input features.



The Bernoulli distribution models the likelihood of binary outcomes $Y \in \{0, 1\}$. The probability for a single observation is:

$P(Y|X) = a(z)^Y \cdot (1 - a(z))^{1-Y}$
where:
 $a(z)$ is the predicted probability of $Y = 1$,
 $1 - a(z)$ is the predicted probability of $Y = 0$


Substituting the Bernoulli probability function, the likelihood becomes:

$$L(W_1, W_2, b) = \prod_{i=1}^n \left[ a(z_i)^{Y_i} \cdot (1 - a(z_i))^{1-Y_i} \right]$$

If the result of this calculation is 0.99, for example, this means that our model is 99% plausible. Conversely, if the result is close to 0, it would mean that our model is highly implausible. Let us apply it to our data
```{r , echo=FALSE, warning=FALSE, message=FALSE}
# Data
X1 <- c(0.5, 1.1, 0.7, 2.0, 2.5, 2.2)
X2 <- c(2.0, 2.1, 2.6, 1.0, 0.7, 0.3)
Y <- c(1, 1, 1, 0, 0, 0)

# Design matrix
X <- cbind(X1, X2, 1)  # Add a column of 1s for the bias term

svd_decomp <- svd(X)
U <- svd_decomp$u
D <- diag(svd_decomp$d)
V <- svd_decomp$v

theta <- V %*% solve(D) %*% t(U) %*% Y

# Extract W1, W2, and b
W1 <- -1
W2 <- 1
b <- 0.1
```
The probability are:
```{r , echo=FALSE, warning=FALSE, message=FALSE}
Z=W1 * X[,1] + W2* X[,2]+ b
sigmo<-function(a) {1/(1+exp(-a))}
s<-data.frame(Probability=sigmo(Z))

s
```

```{r , echo=FALSE, warning=FALSE, message=FALSE}
L= prod(s)
L
library(ggplot2)
x1= c(0.5,1.1,0.7,2,2.5,2.2)
x2=c(2,2.1,2.6,1,0.7,0.3)
y= factor(c(1,1,1,0,0,0), levels=c(0,1))
Y=factor(y)
data= data.frame(cbind(y,x1,x2)  ) 
ggplot(data = data, aes(x = x1, y = x2)) +
  geom_point(aes(colour = Y), size = 3) + 
  scale_colour_manual(values = c("red", "green")) + 
  geom_abline(slope = 1, intercept = 0.1, colour = "black", size = 1.2)+ # Boundary line
  labs(title = "Data separation", x = "x1", y = "x2") + 
  theme_minimal()
```

The likelihood value is `0.003482233`.
In fact, this is the prduct of probability

$L= 0.8320184 * 0.7502601 * 0.8807971 * 0.2890505 * 0.1544653 * 0.1418511$

Let's remark that the likelihood is very low. Can we really deduce that the model is'nt likely?. Imagine that we have more than 10000 lines. The product of all the probability can lead to a number extremely close to 0, at some point, the machine will no longer be able to store the value.
That's why we use the log-likelihood.

**Log-Likelihood Function**

To simplify computation and avoid numerical underflow, we use the log-likelihood, which is the natural logarithm of the likelihood:

$$\ell(W_1, W_2, b) = \log L(W_1, W_2, b)$$

Expanding this, we get:

$$\ell(W_1, W_2, b) = \sum_{i=1}^n \left[ Y_i \cdot \log(a(z_i)) + (1 - Y_i) \cdot \log(1 - a(z_i)) \right]$$

In fact, the log function is monotonically increasing. It therefore preserves order. This means that $a<b$ implies that $log(a)<log(b)$
Meaning that to maximize the likelyhood of the model, we just need to maximize the log-likelihood.
```{r  , echo=FALSE, warning=FALSE, message=FALSE}

library(ggplot2)
library(gridExtra)
# Define the likelihood function and log-likelihood function for a normal distribution
likelihood_function <- function(mu, sigma, x) {
  prod(dnorm(x, mean = mu, sd = sigma))
}

log_likelihood_function <- function(mu, sigma, x) {
  sum(dnorm(x, mean = mu, sd = sigma, log = TRUE))
}

# Sample data (you can replace this with your actual data)
set.seed(123)
x <- rnorm(100, mean = 0, sd = 1)

# Define a range of parameter values to plot
mu_values <- seq(-3, 3, length.out = 100)
sigma_values <- 1  # Fixed standard deviation

# Compute the likelihood and log-likelihood for each value of mu
likelihood_values <- sapply(mu_values, function(mu) likelihood_function(mu, sigma_values, x))
log_likelihood_values <- sapply(mu_values, function(mu) log_likelihood_function(mu, sigma_values, x))

# Create the likelihood plot
likelihood_plot <- ggplot(data.frame(mu = mu_values, value = likelihood_values), aes(x = mu, y = value)) +
  geom_line(color = "blue", size = 1.2) +
  labs(title = "Likelihood Function", x = expression(mu), y = "Likelihood") +
  theme_minimal()

# Create the log-likelihood plot
log_likelihood_plot <- ggplot(data.frame(mu = mu_values, value = log_likelihood_values), aes(x = mu, y = value)) +
  geom_line(color = "red", size = 1.2) +
  labs(title = "Log-Likelihood Function", x = expression(mu), y = "Log-Likelihood") +
  theme_minimal()

# Arrange the plots side by side using gridExtra
grid.arrange(likelihood_plot, log_likelihood_plot, ncol = 2)  
```

Generally, it can be a little bit difficult to maximise a mathematical  function. So by multiplying the log like lihood by (-), our aim will be the minimisation of the loglikelihood function .
maximising $f(x)$ is equal to minimise $-f(x)$



```{r , echo=FALSE, warning=FALSE, message=FALSE}
# Load required library
library(ggplot2)

# Define the log-likelihood function for a normal distribution
log_likelihood_function <- function(mu, sigma, x) {
  sum(dnorm(x, mean = mu, sd = sigma, log = TRUE))
}

# Sample data (you can replace this with your actual data)
set.seed(123)
x <- rnorm(100, mean = 0, sd = 1)

# Define a range of parameter values to plot
mu_values <- seq(-3, 3, length.out = 100)
sigma_values <- 1  # Fixed standard deviation

# Compute the log-likelihood and negative log-likelihood for each value of mu
log_likelihood_values <- sapply(mu_values, function(mu) log_likelihood_function(mu, sigma_values, x))
negative_log_likelihood_values <- -log_likelihood_values

# Find the maximum of log-likelihood and the minimum of negative log-likelihood
max_log_likelihood <- max(log_likelihood_values)
max_mu <- mu_values[which.max(log_likelihood_values)]

min_negative_log_likelihood <- min(negative_log_likelihood_values)
min_mu <- mu_values[which.min(negative_log_likelihood_values)]

# Create a data frame for plotting
data <- data.frame(
  mu = mu_values,
  log_likelihood = log_likelihood_values,
  negative_log_likelihood = negative_log_likelihood_values
)

# Create the plot
ggplot(data, aes(x = mu)) +
  geom_line(aes(y = log_likelihood), color = "blue", size = 1.2, linetype = "solid") +
  geom_line(aes(y = negative_log_likelihood), color = "black", size = 1.2, linetype = "dashed") +
  geom_point(aes(x = max_mu, y = max_log_likelihood), color = "blue", size = 4, shape = 19) +
  geom_point(aes(x = min_mu, y = min_negative_log_likelihood), color = "black", size = 4, shape = 19) +
  annotate("text", x = max_mu, y = max_log_likelihood, label = "Max Log-Likelihood", vjust = -1, color = "blue") +
  annotate("text", x = min_mu, y = min_negative_log_likelihood, label = "Min -Log-Likelihood", vjust = 1.5, color = "red") +
  labs(title = "Log-Likelihood and Negative Log-Likelihood Functions",
       x = expression(mu), y = "Value") +
  theme_minimal()
```

We addd also the term 1/N to normalise our answer. This is how the log loss function is gotten.

$$\ell(W_1, W_2, b) = -1/n[\sum_{i=1}^n \left[ Y_i \cdot \log(a(z_i)) + (1 - Y_i) \cdot \log(1 - a(z_i)) \right]]$$



And there you have it, you now know where cost fomction comes from. Now we're going to use the cost funcion to minimise the errors of the model, using the **Gradient descent**.

### **Gradient descent**

Gradient Descent is an optimization algorithm used to minimize a function by iteratively moving towards the minimum value of the function. In machine learning, this function is typically the cost function (or loss function), which measures how well the model is performing. The goal of gradient descent is to find the parameters (e.g., weights in a linear regression model) that minimize this cost function.


Let's be   $∇J(θ) =\ell(W_1, W_2, b)$


The gradient is the vector of partial derivatives of the cost function with respect to each parameter.

It points in the direction of the steepest ascent of the function. To minimize the function, we move in the opposite direction of the gradient. There is a hyperparameter that controls the step size of each iteration, this is the Learning Rate (α):
**Too small**: Slow convergence.

**Too large**: Risk of overshooting the minimum.

Update Rule:

The parameters are updated iteratively using the formula:
$$θ_j = θ_j - α * (∂J(θ)/∂θ_j)$$

This is repeated until convergence (i.e., when the change in the cost function is very small).

Let's define this function for our data and try to visualise the learning with a 3D plot.

```{r , echo=FALSE, warning=FALSE, message=FALSE}

library(ggplot2)
library(gganimate)
library(plotly)
library(dplyr)

# Data
x1 <- c(0.5,1.1,0.7,2,2.5,2.2)
x2 <- c(2,2.1,2.6,1,0.7,0.3)
y <- factor(c(1,1,1,0,0,0), levels=c(0,1))
data <- data.frame(y, x1, x2)
# Sigmoid function
sigmoid <- function(z) {
  return(1 / (1 + exp(-z)))
}

# Logistic loss function
logistic_loss <- function(W, data) {
  z <- W[1] * data$x1 + W[2] * data$x2 + W[3]
  a <- sigmoid(z)
  y <- as.numeric(as.character(data$y))  # Convert factor to numeric
  loss <- -mean(y * log(a) + (1 - y) * log(1 - a))
  return(loss)
}

# Gradient Descent
gradient_descent <- function(data, lr = 0.1, epochs = 50) {
  W <- c(runif(1), runif(1), runif(1))  # Random initialization
  history <- data.frame(iter = integer(), W1 = numeric(), W2 = numeric(), b = numeric(), loss = numeric())
  
  for (i in 1:epochs) {
    z <- W[1] * data$x1 + W[2] * data$x2 + W[3]
    a <- sigmoid(z)
    y <- as.numeric(as.character(data$y))
    
    # Compute gradients
    dW1 <- mean((a - y) * data$x1)
    dW2 <- mean((a - y) * data$x2)
    db  <- mean(a - y)
    
    # Update parameters
    W[1] <- W[1] - lr * dW1
    W[2] <- W[2] - lr * dW2
    W[3] <- W[3] - lr * db
    
    # Store history
    history <- rbind(history, data.frame(iter = i, W1 = W[1], W2 = W[2], b = W[3], loss = logistic_loss(W, data)))
  }
  
  return(history)
}

# Run Gradient Descent
history <- gradient_descent(data, lr = 0.10, epochs = 50)


# Generate 3D Loss Function Surface
W1_seq <- seq(-3, 3, length.out = 30)
W2_seq <- seq(-3, 3, length.out = 30)
grid <- expand.grid(W1 = W1_seq, W2 = W2_seq)
grid$Loss <- apply(grid, 1, function(w) logistic_loss(c(w[1], w[2], 0), data))

# Create 3D Loss Function Plot
loss_plot <- plot_ly(x = grid$W1, y = grid$W2, z = grid$Loss, type = "mesh3d", colorscale = "Viridis") %>%
  layout(scene = list(xaxis = list(title = "W1"),
                      yaxis = list(title = "W2"),
                      zaxis = list(title = "Loss Function")))

# Add Gradient Descent Path
loss_plot <- loss_plot %>%
  add_trace(x = history$W1, y = history$W2, z = history$loss, type = "scatter3d", mode = "lines+markers",
            marker = list(size = 4, color = "red"), line = list(width = 5, color = "red"))

# Show the 3D animated plot
loss_plot

```

In red, we can see all the point that the algo goes through. Don't forget that we want the minimum of ths function. The function seems to be convexe. Let's extract the parameters corresponding to the min lost.

```{r}
min_lost <- history[which.min(history$loss), ]
print(min_lost)
```


Using this process the final line can be adjust like this

```{r}
# Generate Decision Boundaries Over Iterations
animated_plot <- ggplot() +
  geom_point(data = data, aes(x = x1, y = x2, color = y), size = 3) +
  transition_time(history$iter) +
  labs(title = "Iteration: {frame_time}", x = "X1", y = "X2") +
  geom_abline(data = history, aes(intercept = -b/W2, slope = -W1/W2), color = "black")

# Animate
animate(animated_plot, nframes = 50, fps = 10, duration = 15)

```

Now to classify new data, we just have to compute the probability associated using the value of the variable and the parameters  computed. If the probability is greater than 0.5, the new data belongs to the class 1, otherwise, it belongs to the class 0


Thank you for yor attention. I hope it was helpful!!!!!!!!








<button id="back-to-top" onclick="scrollToTop()">⬆️</button>

<script>
  window.onscroll = function() { scrollFunction() };

  function scrollFunction() {
    let button = document.getElementById("back-to-top");
    if (document.body.scrollTop > 200 || document.documentElement.scrollTop > 200) {
      button.style.display = "block";
    } else {
      button.style.display = "none";
    }
  }

  function scrollToTop() {
    window.scrollTo({top: 0, behavior: 'smooth'});
  }
</script>
