---
title: "K-means algo"
---

K -means clustering is one of the most commonly used clustering algorithms for partitioning observations into a set of 𝑘 groups (i.e. 𝑘 clusters), where 𝑘 is pre-speciﬁed by the analyst. k -means, like other clustering algorithms, tries to classify observations into mutually exclusive groups (or clusters), such that observations within the same cluster are as similar as possible (i.e., high intra-class similarity), whereas observations from diﬀerent clusters are as dissimilar as possible (i.e., low inter-class similarity). In k -means clustering, each cluster is represented by its center (i.e, centroid) which corresponds to the mean of the observation values assigned to the cluster.


### Mathematical Formulation
The K-Means algorithm partitions a dataset into 
$k$ clusters by minimizing the variance within each cluster. Given a dataset $𝑋={𝑥_1,𝑥_2,…,𝑥_𝑛}$  with $𝑛$  observations, the objective function of K-Means ( within-cluster sum of squares (WCSS)) is:
$J = \sum_{i=1}^{K} \sum_{x \in C_i} \| x - \mu_i \|^2$
where:

$𝐶_𝑖$ represents the $𝑖^{𝑡ℎ}$ cluster, $𝜇_𝑖$ is the centroid of cluster 𝑖, computed as:
$\mu_i = \frac{1}{|C_i|} \sum_{x \in C_i} x$

$\| x - \mu_i \|^2 = \sum_{j=1}^{d} (x_j - \mu_{ij})^2$  is the Euclidean distance between a point and its cluster centroid, where 
𝑑 is the number of dimensions in the dataset.

### Steps of the  Algorithm

**Step 1: Initialize centroid**
Randomly select k data points as initial centroids.

**Step 2: Assign data points to clusters**
Assign each data point x to the nearest centroid:

$C_i = \{x:∥x−μ_i∥{2} ≤∥x−μ_j∥^{2}for\ all\ j\} $

**Step 3: Update centroids**
Recompute the centroids as the mean of all points in the cluster:
$\mu_i = \frac{1}{|C_i|} \sum_{x \in C_i} x$

**Step 4: Repeat**
Repeat Steps 2 and 3 until the centroids no longer change significantly.


### Exemple
Let's try it with r 
```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Load necessary libraries
library(ggplot2)
library(gganimate)
library(dplyr)

# Generate synthetic data
set.seed(100)
n <- 100
data <- data.frame(
  x = c(rnorm(n, mean = 2, sd = 1), rnorm(n, mean = 5, sd = 1),rnorm(n, mean = 7, sd = 1)),
  y = c(rnorm(n, mean = 2, sd = 1), rnorm(n, mean = 5, sd = 1),rnorm(n, mean = 7, sd = 1))
)

# K-Means Algorithm
kmeans_custom <- function(data, k, max_iters = 10) {
  # Step 1: Initialize centroids randomly
  centroids <- data[sample(1:nrow(data), k), ]
  history <- list()  # To store centroids and cluster assignments at each iteration
  
  for (iter in 1:max_iters) {
    # Step 2: Assign data points to the nearest centroid
    distances <- as.matrix(dist(rbind(data, centroids)))[1:nrow(data), (nrow(data) + 1):(nrow(data) + k)]
    clusters <- apply(distances, 1, which.min)
    
    # Step 3: Update centroids
    new_centroids <- data %>%
      mutate(cluster = clusters) %>%
      group_by(cluster) %>%
      summarize(across(everything(), mean)) %>%
      select(-cluster)
    
    # Store history for animation
    history[[iter]] <- list(
      centroids = centroids,
      clusters = clusters
    )
    
    # Check for convergence
    if (all(centroids == new_centroids)) {
      break
    }
    
    centroids <- new_centroids
  }
  
  return(list(centroids = centroids, history = history))
}

# Run K-Means
k <- 3
result <- kmeans_custom(data, k)

# Prepare data for animation
animation_data <- data.frame()
for (iter in 1:length(result$history)) {
  iter_data <- data %>%
    mutate(
      cluster = result$history[[iter]]$clusters,
      iteration = iter,
      type = "data"  # Label for data points
    )
  centroids_data <- result$history[[iter]]$centroids %>%
    mutate(
      cluster = 1:k,
      iteration = iter,
      type = "centroid"  # Label for centroids
    )
  animation_data <- rbind(animation_data, iter_data, centroids_data)
}


# Create the animation
p <- ggplot(animation_data, aes(x = x, y = y, color = as.factor(cluster), shape = type, size = type)) +
  geom_point() +
  scale_shape_manual(values = c("data" = 19, "centroid" = 17)) +  # Data points as circles, centroids as crosses
  scale_size_manual(values = c("data" = 3, "centroid" = 10)) +    # Larger size for centroids
  scale_color_manual(values = c("1" = "#68228B", "2" = "#00C5CD", "3" = "#006400", "4" = "purple"))+ # Cluster colors
  labs(title = 'K-Means Clustering: Iteration {frame_time}', x = 'x', y = 'y') +
  transition_time(iteration) +
  ease_aes('linear')

# Render the animation
animate(p, nframes = length(result$history), fps = 2, width = 600, height = 400)

```

Now, how the k is choosen?

### Choice of the k

#### Elbow Method
This method evaluates the Within-Cluster Sum of Squares (WCSS) and looks for an "elbow point" in the curve.

$ WCSS = \sum_{i=1}^{K} \sum_{x \in C_i} \| x - \mu_i \|^2$

We plot the WCSS against 𝑘 and we find the point where the curve decrease slows down significantly (the "elbow").The corresponding 𝑘 is the optimal number of clusters.
```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Load necessary libraries
library(ggplot2)

# Generate synthetic data
set.seed(123)
n <- 100
data <- data.frame(
  x = c(rnorm(n, mean = 2, sd = 1), rnorm(n, mean = 6, sd = 1)),
  y = c(rnorm(n, mean = 2, sd = 1), rnorm(n, mean = 6, sd = 1))
)

# Function to compute WCSS for a given k
compute_wcss <- function(data, k) {
  kmeans_result <- kmeans(data, centers = k, nstart = 10)
  return(kmeans_result$tot.withinss)
}

# Elbow Method
elbow_method <- function(data, max_k = 10) {
  wcss_values <- sapply(1:max_k, function(k) compute_wcss(data, k))
  elbow_data <- data.frame(k = 1:max_k, WCSS = wcss_values)
  
  # Plot the elbow curve
  ggplot(elbow_data, aes(x = k, y = WCSS)) +
    geom_line(color = "blue") +
    geom_point(color = "red") +
    labs(title = "Elbow Method", x = "Number of Clusters (k)", y = "WCSS") +
    theme_minimal()
}

# Run the Elbow Method
elbow_method(data)
```

### Silhouette Method


The Silhouette Score measures how similar a data point is to its own cluster compared to other clusters. For a data point x, the Silhouette Score s(x) is:

$S(x) = \frac{b(x)- a(x)}{max(a(x), b(x))} $
,where:
**a(x)**: Average distance between x and all other points in the same cluster.

**b(x)**: Average distance between x and all points in the nearest neighboring cluster.

The optimal **k** is the one that maximizes the average Silhouette Score.

```{r}
# Load necessary libraries
library(cluster)

# Silhouette Method
silhouette_method <- function(data, max_k = 10) {
  avg_silhouette <- numeric(max_k)
  
  for (k in 2:max_k) {
    kmeans_result <- kmeans(data, centers = k, nstart = 10)
    silhouette_score <- silhouette(kmeans_result$cluster, dist(data))
    avg_silhouette[k] <- mean(silhouette_score[, 3])
  }
  
  silhouette_data <- data.frame(k = 1:max_k, Avg_Silhouette = avg_silhouette)
  
  # Plot the Silhouette Scores
  ggplot(silhouette_data, aes(x = k, y = Avg_Silhouette)) +
    geom_line(color = "blue") +
    geom_point(color = "red") +
    labs(title = "Silhouette Method", x = "Number of Clusters (k)", y = "Average Silhouette Score") +
    theme_minimal()
}

# Run the Silhouette Method
silhouette_method(data)
```