---
title: "Hands-on"
---

Let's start with an example.

Let's suppose that we have two kind of plant that we have to divide into two classes: **toxic** plants represented by class **0** and **healthy** plants represented by class **1**. Each plant has his characteristics **X1** and **X2**. The following table show us what it is about:

| Y  |X1 | X2|
|----|---|---|
| 1  |0.5| 2 |
| 1  |1.1|2.1|
| 1  |0.7|2.6|
| 1  |2  | 1 |
| 1  |2.5|0.7|
| 1  |2.2|0.3|

Let' us have a better look with a graphic


```{r , warning=FALSE, message=FALSE}
library(ggplot2)
x1= c(0.5,1.1,0.7,2,2.5,2.2)
x2=c(2,2.1,2.6,1,0.7,0.3)
y= factor(c(1,1,1,0,0,0), levels=c(0,1))
Y=factor(y)
data= data.frame(cbind(y,x1,x2)  ) 
ggplot(data = data, aes(x = x1, y = x2)) +
  geom_point(aes(colour = Y), size = 3) + 
  scale_colour_manual(values = c("red", "green")) +
  theme_minimal()
```

As we can see, we can provide a line that can separate the two kinds of plant. This line is **the decision boundary**. So we're able to develop a model, based on this line to predict the class of a new plant.
```{r , warning=FALSE, message=FALSE }
library(ggplot2)
x1= c(0.5,1.1,0.7,2,2.5,2.2,1.5)
x2=c(2,2.1,2.6,1,0.7,0.3,0.5)
y= factor(c(1,1,1,0,0,0,2), levels=c(0,1,2))
Y=factor(y)
data= data.frame(cbind(y,x1,x2)  ) 
ggplot(data = data, aes(x = x1, y = x2)) +
  geom_point(aes(colour = Y), size = 3) + 
  scale_colour_manual(values = c("red", "green", "blue")) + 
  geom_abline(slope = 1, intercept = 0.1, colour = "black", size = 1.2) + # Boundary line
  labs(title = "Data separation", x = "x1", y = "x2") + 
  theme_minimal()
```


To de velop the model, we need to know the equation of the line that separate accuratly the two plants. To achieve this, we'll develop a linear model by sending **X1** and **X2** to a neuron, adding a coefficient **W**, respectively **W1** and **W2** and what we call the bias **b**. It'll lead us to :

$$  z(X1,X2)= W1*X1 + W2*X2 + b   $$



The boundary line is the value for which $$ Z(X_1,X_2)=0$$   So by posing $Z(X_1,X_2)=0$, we can find the expression of the equation of the boundary line.
To predict a class for a new  plant, we have to  set the parameters $ W_1 $, $ W_2 $  and  $b$  to separe in the best way the two class and after that we will be able to send a new plant according to the sign of the $Z$ computed. 
Now, to improve this model, one of the best thing that we can do is to add probabilities. The further the plant is from the line, the greater the chance that it belongs to this class. In this context, we use an **activation function** that returns a value close to 0 or 1 as we move away from the line. Generally we use the sigmoiid fonction.

The **sigmoid function** is one of the most common activation functions in artificial neural networks, especially in simpler or shallow models. It maps any real-valued number to a range between **0 and 1**, making it particularly useful for binary classification problems.


The **sigmoid function** $a(Z)$ is defined as:

$a(Z) = \frac{1}{1 + e^{-Z}}$



Where:

- $e$ is the base of the natural logarithm (approximately 2.718).
-  $Z$ is the input to the function, which can be any real number.



The graph of the sigmoid function has an **S-shape** (sigmoid curve) and is asymptotic to both 0 and 1 for extreme values of \( x \). 
```{r , warning=FALSE, message=FALSE}
library(ggplot2)
# Define sigmoid function
sigmoid <- function(x) {
  return(1 / (1 + exp(-x)))
}
x_values <- seq(-10, 10, by = 0.1)
y_values <- sigmoid(x_values)
# Create a data frame for plotting
data <- data.frame(x = x_values, y = y_values)

# Plot the sigmoid curve with color differentiation for both the curve and area under it
ggplot(data, aes(x = x, y = y)) +
  geom_line(size = 1.5, color = "black") +  # Black line for sigmoid curve
  geom_ribbon(data = subset(data, x < 5), aes(ymin = y, ymax = 1), fill = "lightblue", alpha = 0.5) +  # Light blue for the left part under and on the curve
  geom_ribbon(data = subset(data, x >=-5 ), aes(ymin = 0, ymax = y), fill = "lightcoral", alpha = 0.5) +  # Light red for the right part under the curve
  labs(title = "Sigmoid Function ", x = "x", y = "Sigmoid(x)") +
  theme_minimal() +
  theme(axis.title = element_text(size = 12), plot.title = element_text(size = 14, hjust = 0.5))
```





